Concurrent Programming with CXL

### 2020-06-26 book/paper.tex
Still working on the fonts

So if model checking does not prove a program correct, why is it
useful?  To answer that question consider, for example, a sorting
algorithm.
Now suppose we create a test program, a model, that tries sorting
\emph{all} lists of up to five numbers chosen from the set \{ 1,
2, 3, 4, 5 \}.  Model checking proves that for those particular
scenarios the sorting algorithm works: the output is a sorted
permutation of the input.  In some sense it is an excellent test:
it will have considered all \emph{corner cases},
\index{corner case}
including lists where all
numbers are the same, lists that are already sorted or reversely
sorted, etc.  If there is a bug in the sorting algorithm, most
likely it would be triggered and the model checker would produce a
scenario that would make it easy to find the source of the bug.
However, if the model checker does not find any bugs, we do not
know for sure that the algorithm works for lists with more than
five numbers or for lists that have values other than the numbers
1 through 5.  Still, we would expect that the likelihood that there
are bugs remaining in the sorting algorithm is small.
Note, however, that it would be easy to write a program
that sorts all lists of up to five numbers correctly but fails to
do so for a list of 6 numbers.  (Hint: simply use an \texttt{if}
statement.)

@@

Like Python, CXL is an imperative,
dynamically typed, and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Syntax: Every statement in CXL must be terminated by a semicolon
(even \texttt{while} statements and \texttt{def} statements).
In Python semicolons are optional and rarely used.
There is no syntactic significance to indentation in CXL.
Correct indentation in CXL is encouraged but not enforced.
CXL only supports basic operator precedence or associativity.
Use parentheses liberally to remove ambiguity.
\item CXL does not (currently) support floating point, iterators, or I/O;
CXL does support \texttt{for} loops and various comprehensions over sets.
\item Python is object-oriented, supporting classes with methods and
inheritance; CXL has objects but does not support classes.  CXL supports
pointers to objects.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

Figure~\ref{fig:triangle} shows a simple example of a CXL program.
(The code for examples in this book can be found in the \texttt{code} folder under
the name listed in the caption of the example.)
The example is sequential and has a method \texttt{triangle} that takes
an integer number as argument.  Each method has a variable called
\textit{result} that eventually contains the result of the
method (there is no \texttt{return} statement in CXL).  The method
also has a variable called \textit{n} containing the value of the
argument.  The \textit{x..y} notation generates a set containing the numbers
from~\textit{x} to~\textit{y} (inclusive).  The last two lines in the program are
the most interesting.
The first assigns to \textit{x} some unspecified value in the range \texttt{0..N}
and the second verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.

@@

(For this to work, make sure \texttt{cxl} is in your command shell's search path.)
Essentially, the $\texttt{choose}($S$)$
\index{choose operator}
operator provides the input to the program by selecting some value from the
set~$S$, while the $\texttt{assert}$ statement checks that the output is
correct.  If the program is correct, the output of CXL is the size of the
``state graph'' (13 states in this case).  If not, CXL also
reports what went wrong, typically by displaying a summary of an execution in
which something went wrong.

@@

\begin{quote}
\begin{itemize}
\item Before you run the program, what do you think will happen?  Is the
program correct in that \textit{count} will always end up being 2?
\item What are the possible values of \textit{count} at the time of the
\texttt{assert} statement?
(You may assume that \texttt{load} and \texttt{store} instructions of the
underlying machine architecture are atomic---in fact they are.)
\end{itemize}
\end{quote}

@@

CXL makes extensive use of dictionaries.
A dictionary maps values, known as \emph{keys}, to values.
The CXL dictionary syntax and properties are a little different from Python.
Unlike Python, any CXL value can be a key, including another
dictionary.
CXL dictionaries are written as
$\mathtt{dict}\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
If \texttt{d} is a dictionary, and \texttt{k} is a key, then the
following expression retrieves the CXL value that $k$ maps to:
\begin{code}
\begin{verbatim}
d k
\end{verbatim}
\end{code}
This is unfamiliar to Python programers, but in CXL square brackets can be used
in the same way as parentheses, so you can express the same thing in the form
that is familiar to Python programmers:
\begin{code}
\begin{verbatim}
d[k]
\end{verbatim}
\end{code}
However, if \texttt{d = dict\{ .count: 3 \}}, then you can write
can write \texttt{d.count} (which has value 3) instead of having to write
\texttt{d[.count]} (although both will work).
Thus using atoms, a dictionary can be made to look much like a Python object.
The meaning of \texttt{d a b c ...} is \texttt{((((d a) b) c) ...)}.

@@

Again, square brackets and parentheses work the same in CXL, so
\texttt{[a, b, c]} (which looks like a Python list)
is the same CXL value as \texttt{(a, b, c)} (which looks like a Python tuple),
which in turn is the same CXL value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
So if \texttt{x == [ False, True ]},
then \texttt{x[0] == False} and \texttt{x[1] == True}, just like in Python.
However, when creating a singleton list, make sure you include the
comma, as in \texttt{[False,]}.

CXL is not an object-oriented language, so objects don't have
built-in methods.  However, CXL does have some powerful operators to
make up for some of that.
For example, dictionaries have two handy unary operators.
If \texttt{d} is a
dictionary, then \texttt{keys~d} (or equivalently \texttt{keys(d)})
returns the set of keys and \texttt{len~d} returns the size of
this set.

@@

\begin{problems}
\item Figure~\ref{fig:incenter} shows an attempt at trying to fix the code of
    Figure~\ref{fig:inc}.  Run it through CXL and see what happens.  Based on
    the error output, describe in English what is wrong with the code by describing,
    in broad steps, how running the program can get into a bad state.
\item What if we moved line~2 of Figure~\ref{fig:incenter} to after the \texttt{if}
    statement (between lines~7 and~8)?  Do you think that would work?  Run it through
    CXL and describe either why it works or why it does not work.
\end{problems}

@@

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CVM bytecode.
Here \texttt{@cs} is a \emph{label},
\index{label}
identifying a location in the CVM bytecode.  The first thing we need to
ensure is that there can never be two processes in the critical section.
This property is called \texttt{mutual exclusion}.
\index{mutual exclusion}
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

CXL in fact supports this.
It has an operator \texttt{atLabel $L$},
\index{atLabel operator}
where $L$
is the atom containing the name of the label (in this case, \texttt{.cs}).
The operator returns a bag (multiset) of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.
Method \texttt{atLabel} only exists for specification purposes---do not
use it in normal code.
The assertion also makes use of the \texttt{nametag()} operator
\index{nametag operator}
that returns the name tag of the current process.
If you run the code through CXL, the assertion should fail because
there is no code yet for safely entering and exiting the critical section.

However, mutual exclusion by itself is easy to ensure.
For example, we could insert the following code to enter the
critical section:
\begin{code}
\begin{verbatim}
while True:
    pass;
;
\end{verbatim}
\end{code}
This code will surely prevent two or more processes from being
at label \texttt{cs} at the same time.
But it does so by preventing \emph{any} process from reaching
the critical section.
We clearly need another property besides mutual exclusion.

@@

\section*{Exercises}
\begin{problems}
\item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
section, mutual exclusion is violated,
the assertion should fail, and a trace should be reported.
Now plug the following code \texttt{while True: pass;;}
just before entering the critical section
in Figure~\ref{fig:cs} and run CXL again.
Mutual exclusion is guaranteed but progress is violated.
CXL should print a short trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports before
and after inserting the code.
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with CXL and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure it out---as
we will find out, it is possible but not easy.
\end{problems}

@@

In 1981, Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
Figures~\ref{fig:upflags} and~\ref{fig:upturn}, and is presented
in Figure~\ref{fig:peterson}.
A process first indicates its interest in entering the critical section
by setting its flag.
It then politely gives way to the other process should it also want to
enter the critical section---if both do so at the same time one will
win because writes to memory in CXL are atomic.
The process continues to be polite, waiting in a \texttt{while} loop
until either the other process is nowhere near the critical section
($\mathit{flag}[1 - \mathit{self}] == \mathtt{False}$) or has given way
($\mathit{turn} == \mathit{self}$).
Running the algorithm with CXL shows that it satisfies both mutual
exclusion and progress.

@@

One candidate for such an invariant is exclusivity itself.
After all, it certainly holds over the initial state.
And as CXL has already determined, exclusivity is an invariant:
it holds over every reachable state.
Unfortunately, it is not an \emph{inductive} invariant.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 at the start of the \texttt{while} loop.
Also imagine that in this particular state $\mathit{turn} = 1$.  Now let
process~1 make a sequence of steps.  Because $\mathit{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, entering a state that is not exclusive.
So exclusivity is an invariant, but not an inductive invariant.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

So we are looking for an inductive invariant that \emph{implies} exclusivity.
In other words, the set of states where the inducive invariant holds
must be a subset of the set of states where there is at most one process in
the critical section.
Let $\mathcal{C}(i) = \mathit{flags}[1 - i] \land
\mathit{turn} = 1 - i$, that is, the condition on the \texttt{while} loop
for process~$i$.
Let predicate $\mathcal{I}_p(i)$ be the following.
if process $i$ is at label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $\mathcal{C}(i)$ does not hold or process $1-i$ is executing after setting
$\mathit{flags}[i]$ but still before setting \textit{turn} to $1-i$.
More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs) \Rightarrow (\lnot \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate)$.
For Peterson's Algorithm, an inductive invariant that works well is
$\mathcal{I}_p(0) \land \mathcal{I}_p(1)$.

@@

A cool anecdote is the following.  When the author of CXL had to teach
Peterson's Algorithm, he refreshed his memory by looking at the Wikipedia
page.  The page claimed that the following predicate is invariant:
if process $i$ is in the critical section, then $\lnot \mathcal{C}(i)$ (i.e.,
$\mathcal{I}_p(i)$ without the disjunct that process $1-i$ is at label \texttt{gate}).
To demonstrate that this predicate is not invariant, you can remove the
disjunct from Figure~\ref{fig:petersonproof} and run it to get a
counterexample.

@@

A method \texttt{m} with argument \texttt{a} is invoked in its
most basic form as follows (assigning the result to~\texttt{r}).
\begin{code}
\begin{verbatim}
r = m a;
\end{verbatim}
\end{code}
That's right, no parentheses are required.  In fact, if you invoke
\texttt{m(a)}, the argument is \texttt{(a)}, which is the same
as \texttt{a}.
If you invoke \texttt{m()}, the argument is \texttt{()},
which is the empty tuple.
If you invoke \texttt{m(a, b)}, the argument is \texttt{(a, b)},
the tuple consisting of values \texttt{a} and \texttt{b}.

You may note that all this looks familiar.  Indeed, the syntax
is the same as that for dictionaries (see Chapter~\ref{ch:cxlmachine}).
Both dictionaries and methods map CXL values to CXL values,
and their syntax is indistinguishable.
If \texttt{f} is either a method or a
dictionary, and \texttt{x} is an arbitrary CXL value, then
\texttt{f x}, \texttt{f(x)}, and \texttt{f[x]} are all
the same expression in CXL.

@@

CXL is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In CXL, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this (as well as some other uses), CXL supports \emph{pointers}
\index{pointer}
to shared variables.
If \texttt{x} is a shared variable, then the expression \texttt{\&x}
 is a pointer to \texttt{x} (also known as the \emph{address}
\index{address}
of \texttt{x}).
Conversely, if \texttt{p} is a pointer to a shared variable, then the
expression \texttt{\^{}p} is the value of the shared variable.
In particular, \texttt{\^{}\&x == x}.
This is similar to how C pointers work.

Figure~\ref{fig:petersonmethods} again shows Peterson's algorithm,
but this time with methods defined to enter and exit the critical
section.
The name \texttt{mutex} is often used to denote a variable or value
that is used for mutual exclusion.
\texttt{P\_mutex} is a method that returns a ``mutex,'' which, in this
case, is a dictionary that contains Peterson's Algorithm's state:
a turn variable and two flags.
Both methods \texttt{P\_enter} and \texttt{P\_exit} take two arguments:
a pointer to a mutex and the process identifier (0 or 1).
\texttt{(\^{}pm).turn} is the value of the \texttt{.turn} key
in the mutex dictionary.  Note that the parentheses in this expression
are needed, as \texttt{\^{}pm.turn} would wrongly evaluate to
\texttt{\^{}(pm.turn)}.

@@

$\mathcal{I}_1$ is an inductive invariant.
To see that invariant $\mathcal{I}_1$ is maintained, note that
\texttt{\^{}}$p$ == \texttt{True} upon entry of \texttt{tas}
(because of the condition on the \texttt{while} loop that the
\texttt{tas} method is invoked in).
So there are two cases:
\begin{enumerate}
\item \texttt{\^{}}$s$ is \texttt{False} upon entry to \texttt{tas}.
Then upon exit \texttt{\^{}}$p$ == \texttt{False} and \texttt{\^{}}$s$ == \texttt{True},
maintaining the invariant.
\item \texttt{\^{}}$s$ is \texttt{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section.
Invariant $\mathcal{I}_2(i)$ is obvious as (i) process $i$ only proceeds to the critical
section if \textit{private}[$i$] == \texttt{False}, and (ii) no other process modifies
\textit{private}[$i$].

@@

We can do this by adding another process that, in a loop,
checks the invariant.  Figure~\ref{fig:tasinv} shows the code.
Method \texttt{checkInvariant()} checks to see if the invariant holds
in a state.  It introduces a new feature of CXL: the ability to have
variables local to a method.  In this case, the process variable \texttt{sum}
is used to compute the number of shared variables that have value
\texttt{False}.
The function is invoked by in a loop by a process that runs alongside
the other processes.
In CXL, \texttt{assert} statements are executed atomically, so the
evaluation of the assertion is not interleaved with the execution
of other processes.
Because CXL tries every possible execution, the process is guaranteed
to find violations of the invariant if it does not hold.

@@

Step 1 is to enumerate all waiting conditions.  In the case of the reader/writer
problem, there are two: a process that wants to read may have to wait for a
writer to leave the critical region, while a process that wants to write may
have to wait until all readers have left the critical section.  If there are $N$
waiting conditions, the SBS technique will use $N+1$ binary semaphores: one for
each waiting condition and an extra one that processes use when they first try
to enter the critical section and do not yet know if they need to wait.
We shall call this the \texttt{mutex}.

An important invariant of the SBS technique is that the sum of the $N+1$ semaphores
must always be 0 or 1, and it should be 0 when a process is accessing the shared
variables and 1 when not.
As it were, these semaphores taken together are emulating a single binary semaphore.
Initially, \texttt{mutex} is~1 and the other $N$ semaphores
are all~0.  To maintain the invariant, each process alternates \texttt{P} and
\texttt{V} operations, starting with \texttt{P(\&mutex)} when it tries to
enter the CCS, and ending with a \texttt{V} operation on one of the semaphores
(yet to be determined).

@@

Either way, it vacates one of the semaphores next.
\texttt{V\_one()} is the baton passing function that selects which of the
binary semaphores to vacate.
Method \texttt{V\_one()} first checks to see if there are any readers or
writers waiting.  If there are readers waiting and there are no writers
in the critical section, it vacates the semaphore associated with
readers waiting.  If there are writers waiing and there are no readers
nor writers in the critical section, it vacates the semaphore associated
with writers waiting.  If both conditions hold, it could vacate either
one---it would not matter.  However, it can only vacate one of the semaphores.
If none of the conditions hold, \texttt{V\_one()} vacates \texttt{mutex}.

Let us return now to \texttt{rlock\_acquire}. In the case that the process,
say $p$,
found there is a writer in the critical section, it vacates
\texttt{mutex} and procures \texttt{r\_sema},
blocking on the semaphore associated
with waiting readers.  When later some other process passes the baton to
$p$ by incrementing \texttt{r\_sema}, $p$ updates the state by decrementing
\texttt{r\_waiting} and incrementing \texttt{r\_entered}.  $p$
then invokes \texttt{V\_one()} one more time.

>>>>

So if model checking does not prove a program correct, why is it
useful?  To answer that question consider, for example, a sorting
algorithm.
Now suppose we create a test program, a model, that tries sorting
\emph{all} lists of up to five numbers chosen from the set \{ 1,
2, 3, 4, 5 \}.  Model checking proves that for those particular
scenarios the sorting algorithm works: the output is a sorted
permutation of the input.  In some sense it is an excellent test:
it will have considered all \emph{corner cases},
\index{corner case}
including lists where all
numbers are the same, lists that are already sorted or reversely
sorted, etc.  If there is a bug in the sorting algorithm, most
likely it would be triggered and the model checker would produce a
scenario that would make it easy to find the source of the bug.
However, if the model checker does not find any bugs, we do not
know for sure that the algorithm works for lists with more than
five numbers or for lists that have values other than the numbers
1 through 5.  Still, we would expect that the likelihood that there
are bugs remaining in the sorting algorithm is small.
Note, however, that it would be easy to write a program
that sorts all lists of up to five numbers correctly but fails to
do so for a list of 6 numbers.  (Hint: simply use an \textbf{if}
statement.)

@@

Like Python, CXL is an imperative,
dynamically typed, and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Syntax: Every statement in CXL must be terminated by a semicolon
(even \textbf{while} statements and \textbf{def} statements).
In Python semicolons are optional and rarely used.
There is no syntactic significance to indentation in CXL.
Correct indentation in CXL is encouraged but not enforced.
CXL only supports basic operator precedence or associativity.
Use parentheses liberally to remove ambiguity.
\item CXL does not (currently) support floating point, iterators, or I/O;
CXL does support \textbf{for} loops and various comprehensions over sets.
\item Python is object-oriented, supporting classes with methods and
inheritance; CXL has objects but does not support classes.  CXL supports
pointers to objects.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

Figure~\ref{fig:triangle} shows a simple example of a CXL program.
(The code for examples in this book can be found in the \texttt{code} folder under
the name listed in the caption of the example.)
The example is sequential and has a method \texttt{triangle} that takes
an integer number as argument.  Each method has a variable called
\textit{result} that eventually contains the result of the
method (there is no \textbf{return} statement in CXL).  The method
also has a variable called \textit{n} containing the value of the
argument.  The \textit{x..y} notation generates a set containing the numbers
from~\textit{x} to~\textit{y} (inclusive).  The last two lines in the program are
the most interesting.
The first assigns to \textit{x} some unspecified value in the range \texttt{0..N}
and the second verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.

@@

(For this to work, make sure \texttt{cxl} is in your command shell's search path.)
Essentially, the $\texttt{choose}($S$)$
\index{choose operator}
operator provides the input to the program by selecting some value from the
set~$S$, while the $\textbf{assert}$ statement checks that the output is
correct.  If the program is correct, the output of CXL is the size of the
``state graph'' (13 states in this case).  If not, CXL also
reports what went wrong, typically by displaying a summary of an execution in
which something went wrong.

@@

\begin{quote}
\begin{itemize}
\item Before you run the program, what do you think will happen?  Is the
program correct in that \textit{count} will always end up being 2?
\item What are the possible values of \textit{count} at the time of the
\textbf{assert} statement?
(You may assume that \texttt{load} and \texttt{store} instructions of the
underlying machine architecture are atomic---in fact they are.)
\end{itemize}
\end{quote}

@@

CXL makes extensive use of dictionaries.
A dictionary maps values, known as \emph{keys}, to values.
The CXL dictionary syntax and properties are a little different from Python.
Unlike Python, any CXL value can be a key, including another
dictionary.
CXL dictionaries are written as
$\mathtt{dict}\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
If $d$ is a dictionary, and $k$ is a key, then the
following expression retrieves the CXL value that $k$ maps to in $d$:
\begin{code}
$d$ $k$
\end{code}
This is unfamiliar to Python programers, but in CXL square brackets can be used
in the same way as parentheses, so you can express the same thing in the form
that is familiar to Python programmers:
\begin{code}
$d$[$k$]
\end{code}
However, if $d$ = \texttt{dict\{ .count: 3 \}}, then you can write
can write $d$.\texttt{count} (which has value 3) instead of having to write
$d$\texttt{[.count]} (although both will work).
Thus using atoms, a dictionary can be made to look much like a Python object.
The meaning of $d$ $a$ $b$ $...$ is ((($d$ $a$) $b$) $...$).

@@

Again, square brackets and parentheses work the same in CXL, so
\texttt{[a, b, c]} (which looks like a Python list)
is the same CXL value as \texttt{(a, b, c)} (which looks like a Python tuple),
which in turn is the same CXL value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
So if $x$ = \texttt{[ False, True ]},
then $x$[0] = \texttt{False} and $x$[1] = \texttt{True}, just like in Python.
However, when creating a singleton list, make sure you include the
comma, as in \texttt{[False,]}.

CXL is not an object-oriented language, so objects don't have
built-in methods.  However, CXL does have some powerful operators to
make up for some of that.
For example, dictionaries have two handy unary operators.
If $d$ is a
dictionary, then \texttt{keys}~$d$ (or equivalently \texttt{keys}($d$)
returns the set of keys and \texttt{len}~$d$ returns the size of
this set.

@@

\begin{problems}
\item Figure~\ref{fig:incenter} shows an attempt at trying to fix the code of
    Figure~\ref{fig:inc}.  Run it through CXL and see what happens.  Based on
    the error output, describe in English what is wrong with the code by describing,
    in broad steps, how running the program can get into a bad state.
\item What if we moved line~2 of Figure~\ref{fig:incenter} to after the \textbf{if}
    statement (between lines~7 and~8)?  Do you think that would work?  Run it through
    CXL and describe either why it works or why it does not work.
\end{problems}

@@

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CVM bytecode.
Here \texttt{@cs} is a \emph{label},
\index{label}
identifying a location in the CVM bytecode.  The first thing we need to
ensure is that there can never be two processes in the critical section.
This property is called \emph{mutual exclusion}.
\index{mutual exclusion}
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

CXL in fact supports this.
It has an operator \texttt{atLabel}~$L$,
\index{atLabel operator}
where $L$
is the atom containing the name of the label (in this case, \texttt{.cs}).
The operator returns a bag (multiset) of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.
Method \texttt{atLabel} only exists for specification purposes---do not
use it in normal code.
The assertion also makes use of the \texttt{nametag()} operator
\index{nametag operator}
that returns the name tag of the current process.
If you run the code through CXL, the assertion should fail because
there is no code yet for safely entering and exiting the critical section.

However, mutual exclusion by itself is easy to ensure.
For example, we could insert the following code to enter the
critical section:
\begin{code}
\begin{verbatim}
while True:
    pass;
;
\end{verbatim}
\end{code}
This code will surely prevent two or more processes from being
at label \texttt{@cs} at the same time.
But it does so by preventing \emph{any} process from reaching
the critical section.
We clearly need another property besides mutual exclusion.

@@

\section*{Exercises}
\begin{problems}
\item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
section, mutual exclusion is violated,
the assertion should fail, and a trace should be reported.
Now plug the following code \textbf{while} \texttt{True}: \textbf{pass};;
just before entering the critical section
in Figure~\ref{fig:cs} and run CXL again.
Mutual exclusion is guaranteed but progress is violated.
CXL should print a short trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports before
and after inserting the code.
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with CXL and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure it out---as
we will find out, it is possible but not easy.
\end{problems}

@@

In 1981, Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
Figures~\ref{fig:upflags} and~\ref{fig:upturn}, and is presented
in Figure~\ref{fig:peterson}.
A process first indicates its interest in entering the critical section
by setting its flag.
It then politely gives way to the other process should it also want to
enter the critical section---if both do so at the same time one will
win because writes to memory in CXL are atomic.
The process continues to be polite, waiting in a \textbf{while} loop
until either the other process is nowhere near the critical section
($\mathit{flag}[1 - \mathit{self}] == \mathtt{False}$) or has given way
($\mathit{turn} == \mathit{self}$).
Running the algorithm with CXL shows that it satisfies both mutual
exclusion and progress.

@@

One candidate for such an invariant is exclusivity itself.
After all, it certainly holds over the initial state.
And as CXL has already determined, exclusivity is an invariant:
it holds over every reachable state.
Unfortunately, it is not an \emph{inductive} invariant.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 at the start of the \textbf{while} loop.
Also imagine that in this particular state $\mathit{turn} = 1$.  Now let
process~1 make a sequence of steps.  Because $\mathit{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, entering a state that is not exclusive.
So exclusivity is an invariant, but not an inductive invariant.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

So we are looking for an inductive invariant that \emph{implies} exclusivity.
In other words, the set of states where the inducive invariant holds
must be a subset of the set of states where there is at most one process in
the critical section.
Let $\mathcal{C}(i) = \mathit{flags}[1 - i] \land
\mathit{turn} = 1 - i$, that is, the condition on the \textbf{while} loop
for process~$i$.
Let predicate $\mathcal{I}_p(i)$ be the following.
if process $i$ is at label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $\mathcal{C}(i)$ does not hold or process $1-i$ is executing after setting
$\mathit{flags}[i]$ but still before setting \textit{turn} to $1-i$.
More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs) \Rightarrow (\lnot \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate)$.
For Peterson's Algorithm, an inductive invariant that works well is
$\mathcal{I}_p(0) \land \mathcal{I}_p(1)$.

@@

A cool anecdote is the following.  When the author of CXL had to teach
Peterson's Algorithm, he refreshed his memory by looking at the Wikipedia
page.  The page claimed that the following predicate is invariant:
if process $i$ is in the critical section, then $\lnot \mathcal{C}(i)$ (i.e.,
$\mathcal{I}_p(i)$ without the disjunct that process $1-i$ is at label \texttt{@gate}).
To demonstrate that this predicate is not invariant, you can remove the
disjunct from Figure~\ref{fig:petersonproof} and run it to get a
counterexample.

@@

A method \texttt{m} with argument $a$ is invoked in its
most basic form as follows (assigning the result to~\texttt{r}).
\begin{code}
$r$ = m $a$;
\end{code}
That's right, no parentheses are required.  In fact, if you invoke
\texttt{m}($a$), the argument is ($a$), which is the same
as $a$.
If you invoke \texttt{m()}, the argument is \texttt{()},
which is the empty tuple.
If you invoke \texttt{m}($a$, $b$), the argument is ($a$, $b$),
the tuple consisting of values $a$ and $a$.

You may note that all this looks familiar.  Indeed, the syntax
is the same as that for dictionaries (see Chapter~\ref{ch:cxlmachine}).
Both dictionaries and methods map CXL values to CXL values,
and their syntax is indistinguishable.
If \texttt{f} is either a method or a
dictionary, and $x$ is an arbitrary CXL value, then
\texttt{f} $x$, \texttt{f}($x$), and \texttt{f}[$x$] are all
the same expression in CXL.

@@

CXL is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In CXL, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this (as well as some other uses), CXL supports \emph{pointers}
\index{pointer}
to shared variables.
If $x$ is a shared variable, then the expression \&$x$
 is a pointer to $x$ (also known as the \emph{address}
\index{address}
of $x$).
Conversely, if $p$ is a pointer to a shared variable, then the
expression \texttt{\^{}}$p$ is the value of the shared variable.
In particular, \texttt{\^{}\&}$x$ == $x$.
This is similar to how C pointers work.

Figure~\ref{fig:petersonmethods} again shows Peterson's algorithm,
but this time with methods defined to enter and exit the critical
section.
The name \textit{mutex} is often used to denote a variable or value
that is used for mutual exclusion.
\texttt{P\_mutex} is a method that returns a ``mutex,'' which, in this
case, is a dictionary that contains Peterson's Algorithm's state:
a turn variable and two flags.
Both methods \texttt{P\_enter} and \texttt{P\_exit} take two arguments:
a pointer to a mutex and the process identifier (0 or 1).
\texttt{(\^{}}\textit{pm}\texttt{.turn} is the value of the \texttt{.turn} key
in the dictionary that \textit{pm} points to. 
Note that the parentheses in this expression
are needed, as \texttt{\^{}}\textit{pm}\texttt{.turn} would wrongly evaluate to
\texttt{\^{}}(\textit{pm}\texttt{.turn}).

@@

$\mathcal{I}_1$ is an inductive invariant.
To see that invariant $\mathcal{I}_1$ is maintained, note that
\texttt{\^{}}$p$ == \texttt{True} upon entry of \texttt{tas}
(because of the condition on the \textbf{while} loop that the
\texttt{tas} method is invoked in).
So there are two cases:
\begin{enumerate}
\item \texttt{\^{}}$s$ is \texttt{False} upon entry to \texttt{tas}.
Then upon exit \texttt{\^{}}$p$ == \texttt{False} and \texttt{\^{}}$s$ == \texttt{True},
maintaining the invariant.
\item \texttt{\^{}}$s$ is \texttt{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section.
Invariant $\mathcal{I}_2(i)$ is obvious as (i) process $i$ only proceeds to the critical
section if \textit{private}[$i$] == \texttt{False}, and (ii) no other process modifies
\textit{private}[$i$].

@@

We can do this by adding another process that, in a loop,
checks the invariant.  Figure~\ref{fig:tasinv} shows the code.
Method \texttt{checkInvariant()} checks to see if the invariant holds
in a state.  It introduces a new feature of CXL: the ability to have
variables local to a method.  In this case, the process variable \texttt{sum}
is used to compute the number of shared variables that have value
\texttt{False}.
The function is invoked by in a loop by a process that runs alongside
the other processes.
In CXL, \textbf{assert} statements are executed atomically, so the
evaluation of the assertion is not interleaved with the execution
of other processes.
Because CXL tries every possible execution, the process is guaranteed
to find violations of the invariant if it does not hold.

@@

Step 1 is to enumerate all waiting conditions.  In the case of the reader/writer
problem, there are two: a process that wants to read may have to wait for a
writer to leave the critical region, while a process that wants to write may
have to wait until all readers have left the critical section.  If there are $N$
waiting conditions, the SBS technique will use $N+1$ binary semaphores: one for
each waiting condition and an extra one that processes use when they first try
to enter the critical section and do not yet know if they need to wait.
We shall call this the \textit{mutex}.

An important invariant of the SBS technique is that the sum of the $N+1$ semaphores
must always be 0 or 1, and it should be 0 when a process is accessing the shared
variables and 1 when not.
As it were, these semaphores taken together are emulating a single binary semaphore.
Initially, \textit{mutex} is~1 and the other $N$ semaphores
are all~0.  To maintain the invariant, each process alternates \texttt{P} and
\texttt{V} operations, starting with \texttt{P(\&mutex)} when it tries to
enter the CCS, and ending with a \texttt{V} operation on one of the semaphores
(yet to be determined).

@@

Either way, it vacates one of the semaphores next.
\texttt{V\_one()} is the baton passing function that selects which of the
binary semaphores to vacate.
Method \texttt{V\_one()} first checks to see if there are any readers or
writers waiting.  If there are readers waiting and there are no writers
in the critical section, it vacates the semaphore associated with
readers waiting.  If there are writers waiing and there are no readers
nor writers in the critical section, it vacates the semaphore associated
with writers waiting.  If both conditions hold, it could vacate either
one---it would not matter.  However, it can only vacate one of the semaphores.
If none of the conditions hold, \texttt{V\_one()} vacates \textit{mutex}.

Let us return now to \texttt{rlock\_acquire}. In the case that the process,
say $p$,
found there is a writer in the critical section, it vacates
\textit{mutex} and procures \texttt{r\_sema},
blocking on the semaphore associated
with waiting readers.  When later some other process passes the baton to
$p$ by incrementing \texttt{r\_sema}, $p$ updates the state by decrementing
\texttt{r\_waiting} and incrementing \texttt{r\_entered}.  $p$
then invokes \texttt{V\_one()} one more time.
